---
title: "Predictors of Lung Adenocarcinoma with Leptomeningeal Metastases: A 2022 Targeted-therapy-assisted molGPA Model"
author: ""
date: ""
output: pdf_document
---

```{r setup, include=FALSE}
library(mice) 
library(survival)
library(readxl)
library(randomForestSRC)

########## Step 1: load data ###########
data <- read_excel("data/MC_data_0830.xls")
##########-------------- done------------------ ###########

```

# Outline


1. Data summary
2. Missing data
3. Cox model
+ 3.1 Univariate
+ 3.2 Multivariate
+ 3.3 2021 molGPA and C-index
4. Random survival forest



\newpage

# Data summary

The total number of patients is `r (N = dim(data)[1])`. (In the previous dataset, we have 94 patients in total).

There are `r (p = dim(data)[2])` columns in the datasets.

The number of variables (exclude patient ID, time and status) is `r (p = dim(data)[2] - 3)`.



```{r eval=TRUE, echo=FALSE, fig.align='center', fig.height=4, fig.width=6}
########## Step 2: sanity check and summary ###########
# sex: 1: male; 2: female
# table(data$Sex)
# 1  2 
# 47 47 
h1 = hist(as.integer(data$Sex), ylim = c(0, 100), xlab = "Sex", xaxt='n', main = "", 
          breaks = 2, col = c("#5F4B8BFF", "#E69A8DFF"), labels = TRUE)
axis(side=1, at=h1$mids, labels = c("Male", "Female"))

# # age
# summary(data$Age)
# # Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
# # 28.00   52.00   61.00   59.28   67.75   78.00 
h2 = hist(data$Age, ylim = c(0, 38), xlab = "Age", xaxt='n', main = "", labels = TRUE, col = "#2C5F2D")
axis(side=1, at=h2$mids, labels = h2$mids+0.5)
# 
# 
# 
# # smoking status: 1.smoke; 0: not smoking
# table(data$Smoke)
# # 0  1 
# # 69 25 
h3 = hist(as.integer(data$Smoke), ylim = c(0, 150), xlab = "Smoking status", xaxt='n', main = "",
          breaks = 2, col = c("#5F4B8BFF", "#E69A8DFF"), labels = TRUE)
axis(side=1, at=h3$mids, labels = c("Non-smoker", "Smoker"))
# 
# 
# 
# # 颅外转移灶: 0无颅外转移, 1伴有颅外转移
# table(data$`Extracranial metastases`)
h4 = hist(as.integer(data$`Extracranial metastases`), ylim = c(0, 150),
          xlab = "Extracranial metastases", xaxt='n', main = "",
          breaks = 2, col = c("#5F4B8BFF", "#E69A8DFF"), labels = TRUE)
axis(side=1, at=h4$mids, labels = c("No", "Yes"))
# 
# 
# 
# #原发肿瘤距脑膜癌时间（月）: 0为以脑膜癌为首发症状
# summary(data[,6]) 
h5 = hist(as.numeric(c(data[,6][[1]])), ylim = c(0, 100), xlab = "Median time from cancer diagnosis to leptomeningealcarcinomatosis",
          main = "",col = "#2C5F2D", labels = TRUE)
# axis(side=1, at=h5$mids, labels = h5$mids)
length(which(data[,6][[1]] == 0))
text(80, 50, "55 out of 173 patients with median time = 0", cex = 1)
# 
# 
# # 原发灶肺癌是否得到控制: （0未控制，1控制）
# # Extracranial metastasis control
# table(data$`Extracranial metastasis control`)
h6= hist(as.integer(data$`Controlled primary tumor`), ylim = c(0, 130),
         xlab = "Extracranial metastasis control\n(Controlled primary tumor)", xaxt='n', main = "",
         breaks = 2, col = c("#5F4B8BFF", "#E69A8DFF"), labels = TRUE)
axis(side=1, at=h6$mids, labels = c("No", "Yes")) 
# 
# 
# # 死亡与否（0存活，1死亡2失访）
# # Death or not
# table(data$`Death or not`)
h6= barplot(table(data$`Death or not`),
            ylim = c(0, 200),
            xlab = "Death status", xaxt='n', main = "",
            col = c("#5F4B8BFF", "#E69A8DFF","#00539CFF"))
axis(side=1, at= h6,labels = c("Survival", "Death", "Lost to follow-up"))
text(x = h6, y = table(data$`Death or not`),
     label =table(data$`Death or not`),
     pos = 3, cex =1, col = "black")
# 
# 
# 
# # 生存时间（月）
# summary(data$`Overall survival `) # NA = 3
hist(as.integer(data$`Overall survival `) , ylim = c(0, 150),
     xlab = "Overall survival ",
     main = "",
     col = "#2C5F2D", labels = TRUE)
text(30, 50, paste(sum(is.na(data$`Overall survival `)),"out of 173 patients have NA"), cex = 1)
# 
# 
# # KPS评分分级
# #1为＜60分（高危），2为60~70（中危）,3为≥80分（低危）
# table(data$KPS)
h7= barplot(table(data$KPS),
            ylim = c(0, 100),
            xlab = "KPS", xaxt='n', main = "",
            col = c("#5F4B8BFF", "#E69A8DFF","#00539CFF"))
text(x = h7, y = table(data$KPS),
     label =table(data$KPS),
     pos = 3, cex =1, col = "black")
axis(side=1, at= h7,labels = c("High risk", "Medium risk", "Low risk"))
# 
# 
# # KPS评分原始分值 !!!!!!!!!????????????
dt = data$`Karnofsky Performance Status`
h = barplot(table(dt,  useNA = "ifany"),
            ylim = c(0, max(table(dt)) + 15),
            xlab = "Oringal Karnofsky Performance Status",
            xaxt='n', main = "",
            col = "#2C5F2D")
text(x = h, y = table(dt,  useNA = "ifany"),
     label =table(dt,  useNA = "ifany"),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = unique(dt)[order(unique(dt))])
# 
# 
# # 是否基因突变
# # 0为无EGFR/ALK突变，1为有基因突变（和后面靶向药物对应，有基因突变就会有靶向药物）
# table(data$`Gene status`)
h8= barplot(table(data$`Gene status`),
            ylim = c(0, 150),
            xlab = "Gene status",
            xaxt='n', main = "",
            col = c("#5F4B8BFF", "#E69A8DFF"))
text(x = h8, y = table(data$`Gene status`),
     label =table(data$`Gene status`),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h8, labels = c("No", "Yes"))

h8 = barplot(table(data$`Gene status`, useNA = "ifany"),
             ylim = c(0, max(table(data$`Gene status`,useNA = "ifany")) + 15),
             xlab = "Gene status",
             xaxt='n', main = "",
             col = c("#5F4B8BFF", "#E69A8DFF","#00539CFF"))
text(x = h8, y = table(data$`Gene status`,useNA = "ifany"),
     label =table(data$`Gene status`,useNA = "ifany"),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h8, labels = c("No", "Yes","NA"))
# 
# 
# # syndrome 
for (col in 13:31){
  dt = c(data[,col])[[1]]
  h = barplot(table(dt),
              ylim = c(0, max(table(dt)) + 15),
              xlab = colnames(data)[col],
              xaxt='n', main = "",
              col = c("#5F4B8BFF", "#E69A8DFF"))
  text(x = h, y = table(dt),
       label =table(dt),
       pos = 3, cex =1, col = "black")
  axis(side=1, at=h, labels = c("No", "Yes"))
}
# 
# # RANO神经功能评分
# # 症状体征总和：国际上对脑膜癌神经功能评判总分（包含年龄、KPS评分，颅外转移与否，脑转移数目，基因改变情况）
# # RANO Neurological Assessment
# summary(data$`RANO Neurological Assessment`) 
hist(as.integer(data$`RANO Neurological Assessment`),
     ylim = c(0, 100),
     xlab = "RANO Neurological Assessment",
     main = "",
     col = "#2C5F2D", labels = TRUE)
# 
# 
# # 颅内压分级
# # 1为<200,2为200~300,3为＞300
# table(data$CSFpressure)
h = barplot(table(data$CSFpressure),
            ylim = c(0, max(table(data$CSFpressure)) + 15),
            xlab = "CSFpressure",
            xaxt='n', main = "",
            col = c("#5F4B8BFF", "#E69A8DFF", "#00539CFF"))
text(x = h, y = table(data$CSFpressure),
     label =table(data$CSFpressure),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = c("<200", "200~300", ">300"))
# 
# 
# # 腰穿压力（mmH2O）!!!!!!!!!????????????
# summary(data$`Lumbar puncture pressure`)
dt = as.numeric(data$`Lumbar puncture pressure`)
h = hist(dt,
         xlab = "Oringal lumbar puncture pressure",
         main = "",
         ylim = c(0, 50),
         col = "#2C5F2D", labels = TRUE)
text(200, 40, paste(sum(is.na(dt)),"out of 173 patients have NA"), cex = 1)
# 
# 
# 
# # 白细胞. 脑脊液白细胞（>6个为白细胞增高）
dt = c(data[,35])[[1]]
# summary(dt)
hist(data$`White blood cell`,
     ylim = c(0, 150),
     xlab = "White blood cell",
     main = "",
     col = "#2C5F2D", labels = TRUE)
text(3000, 40, paste(sum(is.na(dt)),"out of 173 patients have NA"), cex = 1)
# 
# 
# # 蛋白
dt = c(data[,36])[[1]]
# summary(dt)
hist(data$Protein,
     ylim = c(0, 150),
     xlab = "Protein",
     main = "",
     col = "#2C5F2D", labels = TRUE)
text(15, 40, paste(sum(is.na(dt)),"out of 173 patients have NA"), cex = 1)
# 
# 
# 
# # 糖
dt = c(data[,37])[[1]]
# summary(dt)
hist(data$Glucose,
     ylim = c(0, 55),
     xlab = "Glucose",
     main = "",
     col = "#2C5F2D", labels = TRUE)
text(6, 40, paste(sum(is.na(dt)),"out of 173 patients have NA"), cex = 1)
# 
# 
# # 氯化物
dt = c(data[,38])[[1]]
# summary(dt)
hist(data$Chloride,
     ylim = c(0, 55),
     xlab = "Chloride",
     main = "",
     col = "#2C5F2D", labels = TRUE)
text(110, 50, paste(sum(is.na(dt)),"out of 173 patients have NA"), cex = 1)

# 
# # 影像学是否阳性
# # 影像总和：是否阳性(包含脑膜强化、脑转移、脑积水及脊髓受累与否)
# # Image
dt = c(data[,39])[[1]]
h = barplot(table(dt),
            ylim = c(0, max(table(dt)) + 15),
            xlab = "If image is positive",
            xaxt='n', main = "",
            col = c("#5F4B8BFF", "#E69A8DFF"))
text(x = h, y = table(dt),
     label =table(dt),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = c("No", "Yes"))
# 
# 
# # 是否脑膜强化
dt = c(data[,40])[[1]] # NA = 4
# table(dt,  useNA = "ifany")
h = barplot(table(dt,  useNA = "ifany"),
            ylim = c(0, max(table(dt)) + 15),
            xlab = "Enhancement",
            xaxt='n', main = "",
            col = c("#5F4B8BFF", "#E69A8DFF","#00539CFF"))
text(x = h, y = table(dt,  useNA = "ifany"),
     label =table(dt,  useNA = "ifany"),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = c("No", "Yes", "NA"))
# 
# 
# # Brain metastases, No.
dt = c(data[,41])[[1]]
# table(dt,  useNA = "ifany")
h = barplot(table(dt,  useNA = "ifany"),
            ylim = c(0, max(table(dt)) + 15),
            xlab = "No. of Brain metastases",
            xaxt='n', main = "",
            col = "#2C5F2D")
text(x = h, y = table(dt,  useNA = "ifany"),
     label =table(dt,  useNA = "ifany"),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = unique(dt)[order(unique(dt))])
# 
# # 脑转移
# # 0无转移，1有转移
# # brain metastases
dt = c(data[,42])[[1]]
h = barplot(table(dt),
            ylim = c(0, max(table(dt)) + 15),
            xlab = "Brain metastases",
            xaxt='n', main = "",
            col = c("#5F4B8BFF", "#E69A8DFF"))
text(x = h, y = table(dt),
     label =table(dt),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = c("No", "Yes"))
# 
# # 脑积水
# # Hydrocephalus
dt = c(data[,43])[[1]]
h = barplot(table(dt),
            ylim = c(0, max(table(dt)) + 15),
            xlab = "Hydrocephalus",
            xaxt='n', main = "",
            col = c("#5F4B8BFF", "#E69A8DFF"))
text(x = h, y = table(dt),
     label =table(dt),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = c("No", "Yes"))
# 
# # 脊髓脊膜受累
# # Enhancement of the spinal membrane
dt = c(data[,44])[[1]] # NA = 4
h = barplot(table(dt, useNA = "ifany"),
            ylim = c(0, max(table(dt,useNA = "ifany")) + 15),
            xlab = "Enhancement of the spinal membrane",
            xaxt='n', main = "",
            col = c("#5F4B8BFF", "#E69A8DFF","#00539CFF"))
text(x = h, y = table(dt,useNA = "ifany"),
     label =table(dt,useNA = "ifany"),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = c("No", "Yes","NA"))
# 
# # LM之前靶向
# # Targeted drugs before LM
dt = c(data[,45])[[1]]
h = barplot(table(dt),
            ylim = c(0, max(table(dt)) + 15),
            xlab = "Targeted drugs before LM",
            xaxt='n', main = "",
            col = c("#5F4B8BFF", "#E69A8DFF"))
text(x = h, y = table(dt),
     label =table(dt),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = c("No", "Yes"))
# 
# # LM之前化疗
# # Chemotherapy before LM
dt = c(data[,46])[[1]]
h = barplot(table(dt),
            ylim = c(0, max(table(dt)) + 15),
            xlab = "Chemotherapy before LM",
            xaxt='n', main = "",
            col = c("#5F4B8BFF", "#E69A8DFF"))
text(x = h, y = table(dt),
     label =table(dt),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = c("No", "Yes"))
# 
# # LM前贝伐珠单抗
# # Bevacizumab before LM
dt = c(data[,47])[[1]]
h = barplot(table(dt),
            ylim = c(0, max(table(dt)) + 15),
            xlab = "Bevacizumab before LM",
            xaxt='n', main = "",
            col = c("#5F4B8BFF", "#E69A8DFF"))
text(x = h, y = table(dt),
     label =table(dt),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = c("No", "Yes"))
# 
# # LM靶向药物分级
# # Target stage
# # 0无靶向药物，1为1代或二代靶向药物，3为3代靶向药物。对预后影响较大
dt = as.integer(c(data[,48])[[1]])
# table(dt,  useNA = "ifany")
h = barplot(table(dt,  useNA = "ifany"),
            ylim = c(0, max(table(dt)) + 15),
            xlab = "Target stage",
            xaxt='n', main = "",
            col = c("#5F4B8BFF", "#E69A8DFF","#00539CFF","#2D3FD7FF"))
text(x = h, y = table(dt,  useNA = "ifany"),
     label =table(dt,  useNA = "ifany"),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = c("No", "1", "2","3"))
# 
# 
# # Chemotherapy after LM
dt = c(data[,49])[[1]]
h = barplot(table(dt),
            ylim = c(0, max(table(dt)) + 15),
            xlab = "Chemotherapy after LM",
            xaxt='n', main = "",
            col = c("#5F4B8BFF", "#E69A8DFF"))
text(x = h, y = table(dt),
     label =table(dt),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = c("No", "Yes"))
# 
# # Operation after LM
dt = c(data[,50])[[1]]
h = barplot(table(dt),
            ylim = c(0, max(table(dt)) + 15),
            xlab = "Operation after LM",
            xaxt='n', main = "",
            col = c("#5F4B8BFF", "#E69A8DFF"))
text(x = h, y = table(dt),
     label =table(dt),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = c("No", "Yes"))
# 
# # Becacizumab after LM
dt = c(data[,51])[[1]]
h = barplot(table(dt),
            ylim = c(0, max(table(dt)) + 15),
            xlab = "Becacizumab after LM",
            xaxt='n', main = "",
            col = c("#5F4B8BFF", "#E69A8DFF"))
text(x = h, y = table(dt),
     label =table(dt),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = c("No", "Yes"))
# 
# # Intrathecal after LM
dt = c(data[,52])[[1]]
h = barplot(table(dt),
            ylim = c(0, max(table(dt)) + 15),
            xlab = "Intrathecal after LM",
            xaxt='n', main = "",
            col = c("#5F4B8BFF", "#E69A8DFF"))
text(x = h, y = table(dt),
     label =table(dt),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = c("No", "Yes"))
# 
# # Radio therapy after LM
dt = c(data[,53])[[1]]
h = barplot(table(dt),
            ylim = c(0, max(table(dt)) + 15),
            xlab = "Radio therapy after LM",
            xaxt='n', main = "",
            col = c("#5F4B8BFF", "#E69A8DFF"))
text(x = h, y = table(dt),
     label =table(dt),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = c("No", "Yes"))
#
# # Immune checkpoint inhibitors
dt = c(data[,54])[[1]]
h = barplot(table(dt),
            ylim = c(0, max(table(dt)) + 15),
            xlab = " Immune checkpoint inhibitors",
            xaxt='n', main = "",
            col = c("#5F4B8BFF", "#E69A8DFF"))
text(x = h, y = table(dt),
     label =table(dt),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = c("No", "Yes"))
# 
# # Lung-molGPA score
dt = c(data[,55])[[1]]
h = barplot(table(dt,  useNA = "ifany"),
            ylim = c(0, max(table(dt)) + 15),
            xlab = "2017 Lung-molGPA score",
            xaxt='n', main = "",
            col = "#2C5F2D")
text(x = h, y = table(dt,  useNA = "ifany"),
     label =table(dt,  useNA = "ifany"),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = unique(dt)[order(unique(dt))])
# 
# # Lung-molGPA: 
# # 2017年脑转移预测模型分级1（0-1.0），2（1.5-2.0）,3（2.5-3.0）,4(3.5-4.0) 分值越高，越低危
dt = c(data[,56])[[1]]
h = barplot(table(dt,  useNA = "ifany"),
            ylim = c(0, max(table(dt)) + 15),
            xlab = "2017 Lung-molGPA stage\nThe higher, the better",
            xaxt='n', main = "",
            col = "#2C5F2D")
text(x = h, y = table(dt,  useNA = "ifany"),
     label =table(dt,  useNA = "ifany"),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = unique(dt)[order(unique(dt))])
# 
# 
# # Novel molGPA score
dt = c(data[,57])[[1]]
h = barplot(table(dt,  useNA = "ifany"),
            ylim = c(0, max(table(dt)) + 15),
            xlab = "2019 Lung-molGPA score",
            xaxt='n', main = "",
            col = "#2C5F2D")
text(x = h, y = table(dt,  useNA = "ifany"),
     label =table(dt,  useNA = "ifany"),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = unique(dt)[order(unique(dt))])
# 
# # Novel molGPA
dt = c(data[,58])[[1]]
h = barplot(table(dt,  useNA = "ifany"),
            ylim = c(0, max(table(dt)) + 15),
            xlab = "2019 Lung-molGPA stage\nThe higher, the better",
            xaxt='n', main = "",
            col = "#2C5F2D")
text(x = h, y = table(dt,  useNA = "ifany"),
     label =table(dt,  useNA = "ifany"),
     pos = 3, cex =1, col = "black")
axis(side=1, at=h, labels = unique(dt)[order(unique(dt))])
```



\newpage

# Missing data and imputation

| variable name                      | # of missing | Note                                               |
|------------------------------------|--------------|----------------------------------------------------|
| Overall survival time (month)      | 11           | Patients with death status = 2 (lost to follow-up) |
| Lumbar puncture pressure              | 48           |          |
| White_blood_cell     |     48        |          |
| Protein         | 48            |          |
| Glucose     | 48            |          |
| Chloride | 48 | |
| Enhancement                        | 3            |                                        |
| Enhancement of spinal membrane | 4            |                                        |
| Gene | 18 | |

We deleted 11 patients who are lost to follow-up. After deleting the patients, we have 162 patients left.

```{r, echo = FALSE}
############### Remove the patient with lost follow-up ###########
ind = which(c(data[,8])[[1]] == 2)
data_all = data[-ind,]
```

For the other two variables with missing value, `mice` package is used.

```{r, echo = FALSE, message=FALSE, warning=FALSE, results = 'hide'}
############# Imputation for missing values #######
data_all = as.data.frame(data_all)
dat <- as.data.frame(sapply(data_all, as.numeric))
colnames(dat) = c("Patient_ID", 'Sex', "Age", "Smoke","Extracranial_metastases",
                  "Median_time_dignosis_to_LC", 
                  "Extracranial_metastasis_control",
                  "status","time",
                  "KPS_stage","KPS_score","Gene",
                  "Headache", "nausea_vomiting",
                  "Meningeal_irritation", "Mental_changes",
                  "Cognitive_impairment","Epilepsy","Lalopathy","Ataxia",
                  "Cranialnerves","Diplopia","Vision_loss","Hear_loss",
                  "Dysphagia", "Facioplegia","spinal_nerve", "Paralysis",
                  "Paresthdsia", "Pain", "UriAndDef", "RANO_Neurological_Assessment",
                  "CSFpressure", "Lumbar_puncture_pressure", "White_blood_cell",
                  "Protein", "Glucose","Chloride", "Image", "Enhancement",
                  "Brain_metastases_number", "brain_metastases",
                  "Hydrocephalus", "Enhancement_of_spinal_membrane",
                  "Targeted_drugs_before_LM", "Chemotherapy_before_LM",
                  "Bevacizumab_before_LM", "Target_stage", "Chemotherapy",
                  "Operation", "Becacizumab", "Intrathecal",
                  "Radio_therapy","Immune_checkpoint_inhibitors",
                  "Lung_molGPA_score", "Lung_molGPA", "Novel_molGPA_score", "Novel_molGPA")

set.seed(125) 
tempData = mice(dat)
completedData = complete(tempData, 1)

##########-------------- done------------------ ###########
```


# Data transformation

For analysis, we did the following variable transformation:

+ Age: dichotomize age with cutoff 65, age = 1 if > 65 
+ White blook cell: take log
+ Protein: take log
+ Glucose: take log
+ KPS score: minus mean
+ Lumbar puncture pressure: minus mean


```{r, echo = FALSE}
############## standardize or take log of the variables ######
completedData$Sex = ifelse(completedData$Sex == 1, 0, 1)
completedData$Age = ifelse(completedData$Age > 65, 0, 1) 
completedData$White_blood_cell = log(completedData$White_blood_cell)
completedData$Protein = log(completedData$Protein)
completedData$Glucose = log(completedData$Glucose)
completedData$Lumbar_puncture_pressure = completedData$Lumbar_puncture_pressure  - mean(completedData$Lumbar_puncture_pressure)
completedData$KPS_score = completedData$KPS_score - mean(completedData$KPS_score)
```

# Split the data into two sets

Training set: 80% ( = 130)
Test set: 20% ( = 32)

In the following analysis, we will use training set to identify the significant variables and fit the model. 

\newpage

# Univariate cox model

```{r, echo = FALSE, results = 'hide'}
########## Step 3: univariate analysis ###########
data_to_fit = completedData[,-c(1,8,9,55,56,57,58)]
names= colnames(data_to_fit)
data_to_fit[,52] = completedData$time
data_to_fit[,53] = completedData$status
colnames(data_to_fit)[52] = "time"
colnames(data_to_fit)[53] = "status"

set.seed(100)
training_id = sample(1:162, 130)
testing_id = c(1:162)[-training_id]
test_data = data_to_fit[testing_id,]
data_to_fit = data_to_fit[training_id,]

for (col in c(1:51)){
  if (col %in% c(5, 8, 29, 31, 32, 33, 34, 35, 38)){
    fit = coxph(Surv(time, status) ~ data_to_fit[,col], data = data_to_fit)
  }else {
    fit = coxph(Surv(time, status) ~ as.factor(data_to_fit[,col]), data = data_to_fit)
  }
  
  s = summary(fit)
  pVal = s$sctest[3]
  print("------------------------------------------------------------------")
  
  print(paste('P-value for the log-rank test for variable', names[col], "is", pVal))
  print(s)
}
```


\newpage 


# Multivariate cox model

## With significant variables from univariate model:


```{r, echo = FALSE, message=FALSE, results ='hide'}
# cox model with univariate significant variables
fit_sub = coxph(Surv(time, status) ~ Age + as.factor(Extracranial_metastasis_control) + as.factor(KPS_stage)  +  as.factor(Gene) +  RANO_Neurological_Assessment + Chloride + as.factor(Target_stage), data = data_to_fit)
summary(fit_sub)
```


```{r, echo = FALSE, message=FALSE, results ='hide'}
# cox model with univariate significant variables
fit_sub_gene = coxph(Surv(time, status) ~ Age + as.factor(Extracranial_metastasis_control) + as.factor(KPS_stage)  +  as.factor(Gene) +  RANO_Neurological_Assessment + Chloride, data = data_to_fit)
summary(fit_sub_gene)
```


```{r, echo = FALSE, message=FALSE, results ='hide'}
# cox model with univariate significant variables
fit_sub_target = coxph(Surv(time, status) ~ Age + as.factor(Extracranial_metastasis_control) + as.factor(KPS_stage)  +  RANO_Neurological_Assessment + Chloride + as.factor(Target_stage), data = data_to_fit)
summary(fit_sub_target)
```



\newpage


# 2021 molGPA and C-index

```{r, echo = FALSE, results='hide'}
train_2021 = train_2019 = train_2017 = test_2021 = test_2019 = test_2017 = rep(NA, 20)
ind = 0
for (seed in seq(from = 0, to = 1000, by = 10)){
  ind = ind + 1
  
  data_to_fit = completedData[,-c(1,8,9,55,56,57,58)]
  names= colnames(data_to_fit)
  data_to_fit[,52] = completedData$time
  data_to_fit[,53] = completedData$status
  colnames(data_to_fit)[52] = "time"
  colnames(data_to_fit)[53] = "status"
  
  set.seed(seed)
  training_id = sample(1:162, 130)
  testing_id = c(1:162)[-training_id]
  test_data = data_to_fit[testing_id,]
  data_to_fit = data_to_fit[training_id,]
  
  
  ########### Training ##########
  v0_2021 = ifelse(data_to_fit$Gene == 0, 0, 0.5)
  
  # control
  v1_2021 = ifelse(data_to_fit$Extracranial_metastasis_control == 1, 0.5, 0)
  
  # KPS_score
  ind1 = which(data_to_fit$KPS_score < 60)
  ind2 = which(60 <= data_to_fit$KPS_score & data_to_fit$KPS_score <= 70)
  ind3 = which(80 <= data_to_fit$KPS_score & data_to_fit$KPS_score <= 100)
  v2_2021 = rep(NA, 130)
  v2_2021[ind1] = 0
  v2_2021[ind2] = 0.5
  v2_2021[ind3] = 1
  
  # RANO
  ind1 = which(data_to_fit$RANO_Neurological_Assessment > 5.75)
  ind2 = which(2 < data_to_fit$RANO_Neurological_Assessment & data_to_fit$RANO_Neurological_Assessment <= 5.75)
  ind3 = which(data_to_fit$RANO_Neurological_Assessment <= 2)
  v3_2021 = rep(NA, 130)
  v3_2021[ind1] = 0
  v3_2021[ind2] = 0.5
  v3_2021[ind3] = 1
  
  # target 
  ind1 = which(data_to_fit$Target_stage == 0)
  ind2 = which(data_to_fit$Target_stage == 1 | data_to_fit$Target_stage == 2)
  ind3 = which(data_to_fit$Target_stage == 3)
  v4_2021 = rep(NA, 130)
  v4_2021[ind1] = 0
  v4_2021[ind2] = 0.5
  v4_2021[ind3] = 1
  
  
  molGPA_2021 =  v0_2021+v1_2021 + v2_2021 + v3_2021 + v4_2021
  
  
  c_index_2021 = concordance(Surv(time, status) ~ molGPA_2021, data = data_to_fit)
  c_index_2017 = concordance(Surv(time, status) ~ completedData[training_id,]$Lung_molGPA_score, data = completedData[training_id,])
  c_index_2019 = concordance(Surv(time, status) ~ completedData[training_id,]$Novel_molGPA_score, data = completedData[training_id,])
  
  train_2021[ind] = c_index_2021$concordance
  train_2019[ind] = c_index_2019$concordance
  train_2017[ind] = c_index_2017$concordance
  
  ########### Test ##########
  v0_2021 = ifelse(test_data$Gene == 0, 0, 0.5)
  
  # control
  v1_2021 = ifelse(test_data$Extracranial_metastasis_control == 1, 0.5, 0)
  
  # KPS_score
  ind1 = which(test_data$KPS_score < 60)
  ind2 = which(60 <= test_data$KPS_score & test_data$KPS_score <= 70)
  ind3 = which(80 <= test_data$KPS_score & test_data$KPS_score <= 100)
  v2_2021 = rep(NA, 32)
  v2_2021[ind1] = 0
  v2_2021[ind2] = 0.5
  v2_2021[ind3] = 1
  
  # RANO
  ind1 = which(test_data$RANO_Neurological_Assessment > 5.75)
  ind2 = which(2 < test_data$RANO_Neurological_Assessment & test_data$RANO_Neurological_Assessment <= 5.75)
  ind3 = which(test_data$RANO_Neurological_Assessment <= 2)
  v3_2021 = rep(NA, 32)
  v3_2021[ind1] = 0
  v3_2021[ind2] = 0.5
  v3_2021[ind3] = 1
  
  # target 
  ind1 = which(test_data$Target_stage == 0)
  ind2 = which(test_data$Target_stage == 1 | test_data$Target_stage == 2)
  ind3 = which(test_data$Target_stage == 3)
  v4_2021 = rep(NA, 32)
  v4_2021[ind1] = 0
  v4_2021[ind2] = 0.5
  v4_2021[ind3] = 1
  
  
  molGPA_2021_test =  v0_2021+v1_2021 + v2_2021 + v3_2021 + v4_2021
  
  
  c_index_2021_test = concordance(Surv(time, status) ~ molGPA_2021_test, data = test_data)
  c_index_2017_test = concordance(Surv(time, status) ~ completedData[testing_id,]$Lung_molGPA_score, data = completedData[testing_id,])
  c_index_2019_test = concordance(Surv(time, status) ~ completedData[testing_id,]$Novel_molGPA_score, data = completedData[testing_id,])
  
  
  test_2021[ind] = c_index_2021_test$concordance
  test_2019[ind] = c_index_2019_test$concordance
  test_2017[ind] = c_index_2017_test$concordance
}
mean(train_2021); round(quantile(train_2021, c(0.025, 0.975)),3)
mean(train_2019); round(quantile(train_2019, c(0.025, 0.975)),3)
mean(train_2017); round(quantile(train_2017, c(0.025, 0.975)),3)
mean(test_2021); round(quantile(test_2021, c(0.025, 0.975)),3)
mean(test_2019); round(quantile(test_2019, c(0.025, 0.975)),3)
mean(test_2017); round(quantile(test_2017, c(0.025, 0.975)),3)
```



\newpage

##  Kaplan-Meier curves for overall survival probability prediction of all 162 patients by group. 

Group 1 (high risk group): molGPA_2021 = 0

Group 2 (mediate risk group): molGPA_2021 = 0.5 or 1

Group 3 (low risk group): molGPA_2021 >= 1.5

```{r, echo = FALSE, message=FALSE, eval=FALSE}
data_all = as.data.frame(data_all)
dat <- as.data.frame(sapply(data_all, as.numeric))
colnames(dat) = c("Patient_ID", 'Sex', "Age", "Smoke","Extracranial_metastases",
                  "Median_time_dignosis_to_LC",
                  "Extracranial_metastasis_control",
                  "status","time",
                  "KPS_stage","KPS_score","Gene",
                  "Headache", "nausea_vomiting",
                  "Meningeal_irritation", "Mental_changes",
                  "Cognitive_impairment","Epilepsy","Lalopathy","Ataxia",
                  "Cranialnerves","Diplopia","Vision_loss","Hear_loss",
                  "Dysphagia", "Facioplegia","spinal_nerve", "Paralysis",
                  "Paresthdsia", "Pain", "UriAndDef", "RANO_Neurological_Assessment",
                  "CSFpressure", "Lumbar_puncture_pressure", "White_blood_cell",
                  "Protein", "Glucose","Chloride", "Image", "Enhancement",
                  "Brain_metastases_number", "brain_metastases",
                  "Hydrocephalus", "Enhancement_of_spinal_membrane",
                  "Targeted_drugs_before_LM", "Chemotherapy_before_LM",
                  "Bevacizumab_before_LM", "Target_stage", "Chemotherapy",
                  "Operation", "Becacizumab", "Intrathecal",
                  "Radio_therapy","Immune_checkpoint_inhibitors",
                  "Lung_molGPA_score", "Lung_molGPA", "Novel_molGPA_score", "Novel_molGPA")
set.seed(100)
tempData = mice(dat)
completedData = complete(tempData, 1)
completedData$Sex = ifelse(completedData$Sex == 1, 0, 1)
data_to_fit = completedData[,-c(1,8,9,55,56,57,58)]
names= colnames(data_to_fit)
data_to_fit[,52] = completedData$time
data_to_fit[,53] = completedData$status
colnames(data_to_fit)[52] = "time"
colnames(data_to_fit)[53] = "status"

########### all ##########
# control
v1_2021 = ifelse(data_to_fit$Extracranial_metastasis_control == 1, 0.5, 0)

# KPS_score
ind1 = which(data_to_fit$KPS_score < 60)
ind2 = which(60 <= data_to_fit$KPS_score & data_to_fit$KPS_score <= 70)
ind3 = which(80 <= data_to_fit$KPS_score & data_to_fit$KPS_score <= 100)
v2_2021 = rep(NA, 130)
v2_2021[ind1] = 0
v2_2021[ind2] = 0.5
v2_2021[ind3] = 1

# RANO
ind1 = which(data_to_fit$RANO_Neurological_Assessment > 5.75)
ind2 = which(2 < data_to_fit$RANO_Neurological_Assessment & data_to_fit$RANO_Neurological_Assessment <= 5.75)
ind3 = which(data_to_fit$RANO_Neurological_Assessment <= 2)
v3_2021 = rep(NA, 130)
v3_2021[ind1] = 0
v3_2021[ind2] = 0.5
v3_2021[ind3] = 1

# target
ind1 = which(data_to_fit$Target_stage == 0)
ind2 = which(data_to_fit$Target_stage == 1 | data_to_fit$Target_stage == 2)
ind3 = which(data_to_fit$Target_stage == 3)
v4_2021 = rep(NA, 130)
v4_2021[ind1] = 0
v4_2021[ind2] = 0.5
v4_2021[ind3] = 1


molGPA_2021 =  v1_2021 + v2_2021 + v3_2021 + v4_2021

group1_ind = which(molGPA_2021 == 0)
group2_ind = which(molGPA_2021 == 0.5 | molGPA_2021 == 1)
group3_ind = which(molGPA_2021 >= 1.5)
data_to_fit$group = rep(NA, dim(data_to_fit)[1])
data_to_fit$group[group1_ind] = 1
data_to_fit$group[group2_ind] = 2
data_to_fit$group[group3_ind] = 3
train_fit = survfit(Surv(time, status) ~ group, data = data_to_fit) # observed
plot(train_fit, xlab="Time (Months)", ylab="Survival Probability",main="KM Survival Curves",
     lty=1, col = c(2,3,4))
legend("topright", inset=.05, c("Group 1 (high risk, GPA = 0)",
                                "Group 2 (mediate risk, GPA = 0.5-1)",
                                "Group 3 (low risk, GPA > 1.5)",
                                "Observed",
                                "Predicted"),
       col = c(2,3,4,1,1), lty=c(1,1,1,1,2))

# predicted
cum_hazard_1 = basehaz(fit_sub_1)$hazard
cum_hazard_2 = basehaz(fit_sub_2)$hazard
cum_hazard_3 = basehaz(fit_sub_3)$hazard
lines( basehaz(fit_sub_1)$time, exp(-cum_hazard_1), col = 2, lty = 2)
lines( basehaz(fit_sub_2)$time, exp(-cum_hazard_2), col = 3, lty = 2)
lines( basehaz(fit_sub_3)$time, exp(-cum_hazard_3), col = 4, lty = 2)
```



\newpage

# Random survival forest

## Significant varibles:

Note: no dichotomization or standardization


```{r, eval = FALSE, echo = FALSE}
########## randomsurvivalforest/randomforest (RSF/RF) ######
data_all = as.data.frame(data_all)
dat <- as.data.frame(sapply(data_all, as.numeric))
colnames(dat) = c("Patient_ID", 'Sex', "Age", "Smoke","Extracranial_metastases",
                  "Median_time_dignosis_to_LC", 
                  "Extracranial_metastasis_control",
                  "status","time",
                  "KPS_stage","KPS_score","Gene",
                  "Headache", "nausea_vomiting",
                  "Meningeal_irritation", "Mental_changes",
                  "Cognitive_impairment","Epilepsy","Lalopathy","Ataxia",
                  "Cranialnerves","Diplopia","Vision_loss","Hear_loss",
                  "Dysphagia", "Facioplegia","spinal_nerve", "Paralysis",
                  "Paresthdsia", "Pain", "UriAndDef", "RANO_Neurological_Assessment",
                  "CSFpressure", "Lumbar_puncture_pressure", "White_blood_cell",
                  "Protein", "Glucose","Chloride", "Image", "Enhancement",
                  "Brain_metastases_number", "brain_metastases",
                  "Hydrocephalus", "Enhancement_of_spinal_membrane",
                  "Targeted_drugs_before_LM", "Chemotherapy_before_LM",
                  "Bevacizumab_before_LM", "Target_stage", "Chemotherapy",
                  "Operation", "Becacizumab", "Intrathecal",
                  "Radio_therapy","Immune_checkpoint_inhibitors",
                  "Lung_molGPA_score", "Lung_molGPA", "Novel_molGPA_score", "Novel_molGPA")
set.seed(100)
tempData = mice(dat)
completedData = complete(tempData, 1)
completedData$Sex = ifelse(completedData$Sex == 1, 0, 1)
data_to_fit = completedData[,-c(1,8,9,55,56,57,58)]
names= colnames(data_to_fit)
data_to_fit[,52] = completedData$time
data_to_fit[,53] = completedData$status
colnames(data_to_fit)[52] = "time"
colnames(data_to_fit)[53] = "status"

set.seed(100)
training_id = sample(1:162, 130)
testing_id = c(1:162)[-training_id]
test_data = data_to_fit[testing_id,]
data_to_fit = data_to_fit[training_id,]

set.seed(100)
fit_RSF <- rfsrc(Surv(time, status) ~ as.factor(Sex) +
                   Age +  as.factor(Smoke) +
                   as.factor(Extracranial_metastases) +
                   Median_time_dignosis_to_LC +
                   as.factor(Extracranial_metastasis_control) +
                   as.factor(KPS_stage) +
                   KPS_score +
                   as.factor(Gene) +
                   as.factor(Headache) +
                   as.factor(nausea_vomiting) +
                   as.factor(Meningeal_irritation) +
                   as.factor(Mental_changes) +
                   as.factor(Cognitive_impairment) +
                   as.factor(Epilepsy) +
                   as.factor(Lalopathy) +
                   as.factor(Ataxia) +
                   as.factor(Cranialnerves) +
                   as.factor(Diplopia) +
                   as.factor(Vision_loss) +
                   as.factor(Hear_loss) +
                   as.factor(Dysphagia) +
                   as.factor(Facioplegia) +
                   as.factor(spinal_nerve) +
                   as.factor(Paralysis) +
                   as.factor(Paresthdsia) +
                   as.factor(Pain) +
                   as.factor(UriAndDef) +
                   RANO_Neurological_Assessment +
                   as.factor(CSFpressure) +
                   Lumbar_puncture_pressure +
                   White_blood_cell +
                   Protein +
                   Glucose +
                   Chloride +
                   as.factor(Image) +
                   as.factor(Enhancement) +
                   Brain_metastases_number +
                   as.factor(brain_metastases) +
                   as.factor(Hydrocephalus) +
                   as.factor(Enhancement_of_spinal_membrane) +
                   as.factor(Targeted_drugs_before_LM) +
                   as.factor(Chemotherapy_before_LM) +
                   as.factor(Bevacizumab_before_LM) +
                   as.factor(Target_stage) +
                   as.factor(Chemotherapy) +
                   as.factor(Operation) +
                   as.factor(Becacizumab) +
                   as.factor(Intrathecal) +
                   as.factor(Radio_therapy) + 
                   as.factor(Immune_checkpoint_inhibitors), 
                 data = data_to_fit,importance = TRUE, block.size=1)

predict = predict.rfsrc(fit_RSF, newdata = test_data)


##########-------------- done------------------ ###########
```



## Top variables
```{r echo=FALSE, eval = FALSE}
set.seed(125)
fit_RSF_sub_Stage <- rfsrc(Surv(time, status) ~ as.factor(KPS_stage) + 
                             RANO_Neurological_Assessment +
                             as.factor(Target_stage) + 
                             Chloride + 
                             as.factor(Extracranial_metastasis_control)  +
                             Age + Gene + as.factor(Chemotherapy_before_LM) + Median_time_dignosis_to_LC +
                             Brain_metastases_number + Glucose, 
                           data = data_to_fit,importance = TRUE, block.size=1)
plot(fit_RSF_sub_Stage)

## added by Jessie on 05/05/2022
# To create a new figure suggested by reviewer 1
jk.obj <- subsample(fit_RSF_sub_Stage)
par(oma = c(0.5, 10, 0.5, 0.5))
par(cex.axis = 1.0, cex.lab = 1.0, cex.main = 1.0, mar = c(6.0,17,1,1), mgp = c(4, 1, 0))
plot(jk.obj, xlab = "Variable Importance (x 100)", cex = 1.2)

predict_sub_stage = predict.rfsrc(fit_RSF_sub_Stage, newdata = test_data)

c_index_list_test = c_index_list_training = c()
ind = 0
for (seed in seq(from = 0, to = 1000, by = 10)){
  print(ind)
  ind = ind + 1
  
  data_to_fit = completedData[,-c(1,8,9,55,56,57,58)]
  names= colnames(data_to_fit)
  data_to_fit[,52] = completedData$time
  data_to_fit[,53] = completedData$status
  colnames(data_to_fit)[52] = "time"
  colnames(data_to_fit)[53] = "status"
  
  set.seed(seed)
  training_id = sample(1:162, 130)
  testing_id = c(1:162)[-training_id]
  test_data = data_to_fit[testing_id,]
  data_to_fit = data_to_fit[training_id,]
  

  fit_RSF_sub_Stage <- rfsrc(Surv(time, status) ~ as.factor(KPS_stage) + 

                               RANO_Neurological_Assessment +
                               as.factor(Target_stage) + 
                               Chloride + 
                               as.factor(Extracranial_metastasis_control)  +
                               Age + Gene + as.factor(Chemotherapy_before_LM) + Median_time_dignosis_to_LC +
                               Brain_metastases_number + Glucose, 

  
  c_index_list_training[ind] = fit_RSF_sub_Stage$err.rate[500]
  
  predict_sub_stage = predict.rfsrc(fit_RSF_sub_Stage, newdata = test_data)
  c_index_list_test[ind] = predict_sub_stage$err.rate[500]
}
quantile(1- c_index_list_training, c(0.025, 0.5, 0.975))
quantile(1 - c_index_list_test, c(0.025, 0.5, 0.975))



```


